Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-30 WholePlotTrajectories.tex
Update WholePlotTrajectories.tex

The taxonomic differences among communities, marked before disturbance
by the distinct starting points on the NMDS axis 2, were maintained
throughout recovery trajectories. More than commonly thought,
post-disturbance trajectories depended on community initial composition,
that partly determined the pool of recruited species and constrained the
trajectories towards the initial composition. The high resilience of
communities taxonomy revealed that species not belonging to the
pre-disturbance community were hardly recruited because of the
commonness of dispersal limitation among tropical tree species
\citep{Svenning2005}.

Conversely, disturbed communities followed functional trajectories
similar in terms of functional composition. As pre-disturbance surviving
trees mirror the initial community \citep{Herault2018}, changes in
functional composition relied upon the recruitment of species or
functional types that were infrequent or absent before disturbance.
Competitive pioneers became dominant in filling the environmental niches
of high availability of light, space and nutrient vacated by the
disturbance. The recruitment of pioneers changed community functional
composition in the same way for all disturbance intensity towards more
resource-acquisitive strategies, moving community functional composition
right along the first axis in Figure \ref{fig:NMDSplans}
\citep{Westoby1998, Wright2004, Reich2014}. Thereafter long-lived, more
resistant and shade-tolerant species excluded the first established
pioneers and started the recovery of pre-disturbance functional
composition, moving similarly community functional composition left
along the first axis and upward along the second axis in Figure
\ref{fig:NMDSplans}.

@@

The IDH predicted well the disturbance impact on community taxonomic
richness, enhanced until an intensity threshold (20-25\% AGB loss), and
to some extent on taxonomic evenness, somewhat decoupled from the
disturbance intensity as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. The disturbance intensity determined the balance in
the community between pre-disturbance surviving trees and those
recruited afterward. The pool of true pioneer species specifically
recruited after disturbance is restricted in the Guiana Shield to a few
common genera (e.g.~Cecropia spp., Vismia spp.) \citep{Guitet2018}.
Below the intensity threshold the size of the surviving community
maintained the pre-disturbance high taxonomic richness while the
recruitment of pioneers, infrequent or absent before disturbance,
increased both community taxonomic richness and evenness. Beyond the
intensity threshold, the disturbance decreased the taxonomic richness of
surviving trees which was not offset by the enrichment of pioneers, so
that the overall community taxonomic richness decreased according to the
disturbance intensity \citep{Molino2001}. For community taxonomic
evenness the disturbance impact was similar but slighter, as the
evenness is less sensitive to the loss of rare species. Taxonomic
evenness rather represented the increasing dominance of pionneers that
balanced the usual hyper-dominance of tropical forest below the
intensity threshold, thus increasing community overall evenness up to
the intensity threshold beyond which pioneers became in turn highly
dominant and decreased the overall evenness \citep{Baraloto2012a}.

@@

For 15 years the species loss during disturbance, determined by the
disturbance intensity, commensurately decreased the functional
redundancy within the pre-disturbance functional space. The redundancy
decrease was not compensated in the first place as the first recruited
pionneers were functionally different from the pre-disturbance
functional composition. Progressively though, first established species
were replaced by more competitive long-lived pionneers or
late-successional species resembling more the pre-disturbance functional
composition and restoring the functional redundancy. This replacement
was stochastic and followed the lottery recruitment rules, implying a
recruitment eased for the first recruited species but then increasingly
hampered by the emergence of interspecific competition
\citep{Busing2002}. Along time the recovery of infrequent species was
increasingly difficult, so that the time for the full recovery of the
functional redundancy, in some communities just initiated after 30
years, was difficult to estimate \citep{Elmqvist2003, Diaz2005}.

@@

Our study revealed community recovery through the combination of
deterministic processes driving their convergence in functional
composition, and dispersal limitation maintaining their divergence in
taxonomic composition. The IDH was validated for community taxonomic
richness and, to some extent, taxonomic evenness but disproved regarding
community functional richness and evennes that were enhanced for any
disturbance intensity by the high functional differences of pioneers
compared to late-successional functoinal composition. The IDH was
translated in time by the recruitment, beyond an intensity threshold, of
short-lived pionneers that prevented in the first times the
establishment of more diverse long-lived pionneers, recruited otherwise
below the intensity threshold. The resilience of tropical forests proved
consistent although several decades-long. Still, the disturbance impact
on communities redundancy cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Gourlet-Fleury2005}.

>>>>

The taxonomic differences among local communities, marked before disturbance
by the distinct starting points on the NMDS axis 2, were maintained
throughout recovery trajectories. More than commonly thought,
post-disturbance trajectories depended on community initial composition,
that partly determined the pool of recruited species and constrained the
trajectories towards the initial composition. The high resilience of
communities taxonomy revealed that species not belonging to the
pre-disturbance community were hardly recruited because of the
commonness of dispersal limitation among tropical tree species
\citep{Svenning2005}.

Conversely, disturbed communities followed functional trajectories
that are highly similar in terms of functional composition. As pre-disturbance surviving
trees mirror the initial community \citep{Herault2018}, changes in
functional composition relied upon the recruitment of species or
functional types that were infrequent or absent before disturbance.
Competitive pioneers became dominant in filling the environmental niches
of high availability of light, space and nutrients vacated by the
disturbance. The recruitment of pioneers changed community functional
composition in the same way for all disturbance intensity towards more
resource-acquisitive strategies, moving community functional composition
right along the first axis in Figure \ref{fig:NMDSplans}
\citep{Westoby1998, Wright2004, Reich2014}. Thereafter long-lived, more
resistant and shade-tolerant species excluded the first established
pioneers and started the recovery of pre-disturbance functional
composition, moving similarly community functional composition left
along the first axis and upward along the second axis in Figure
\ref{fig:NMDSplans}.

@@

The IDH well predicted the disturbance impact on community taxonomic
richness, enhanced until an intensity threshold (20-25\% AGB loss), and
to some extent on taxonomic evenness, somewhat decoupled from the
disturbance intensity as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. The disturbance intensity determined the balance in
the community between pre-disturbance surviving trees and those
recruited afterward. The pool of true pioneer species specifically
recruited after disturbance is restricted in the Guiana Shield to a few
common genera (e.g.~Cecropia spp., Vismia spp.) \citep{Guitet2018}.
Below the intensity threshold the size of the surviving community
maintained the pre-disturbance high taxonomic richness while the
recruitment of pioneers, infrequent or absent before disturbance,
increased both community taxonomic richness and evenness. Beyond the
intensity threshold, the disturbance decreased the taxonomic richness of
surviving trees which was not offset by the enrichment of pioneers, so
that the overall community taxonomic richness decreased according to the
disturbance intensity \citep{Molino2001}. For community taxonomic
evenness the disturbance impact was similar but slighter, as the
evenness is less sensitive to the loss of rare species. Taxonomic
evenness rather represented the increasing dominance of pionneers that
balanced the usual hyper-dominance of a few species in tropical forests below the
intensity threshold, thus increasing community overall evenness up to
the intensity threshold beyond which pioneers became in turn highly
dominant and decreased the overall evenness \citep{Baraloto2012a}.

@@

For 15 years the species loss during disturbance, determined by the
disturbance intensity, commensurately decreased the functional
redundancy within the pre-disturbance functional space. The redundancy
decrease was not compensated in the first place as the first recruited
pionneers were functionally different from the pre-disturbance
functional composition. Progressively though, first established species
were replaced by more competitive long-lived pionneers or
late-successional species resembling more the pre-disturbance functional
composition and restoring the functional redundancy. This replacement
was stochastic and followed the lottery recruitment rules, implying a
recruitment eased for the first recruited species but then increasingly
hampered by the emergence of interspecific competition
\citep{Busing2002}. Along time the recovery of infrequent species was
increasingly difficult, so that the time for the full recovery of the
functional redundancy, in some communities just initiated after 30
years, was extremely difficult to estimate \citep{Elmqvist2003, Diaz2005}.

@@

Our study revealed community recovery through the combination of
deterministic processes driving their convergence in functional
composition, and dispersal limitation maintaining their divergence in
taxonomic composition. The IDH was validated for community taxonomic
richness and, to some extent, taxonomic evenness but disproved regarding
community functional richness and evennes that were enhanced for any
disturbance intensity by the high functional differences of pioneers
compared to late-successional functoinal composition. The IDH was
translated in time by the recruitment, beyond an intensity threshold, of
short-lived pionneers that prevented in the first times the
establishment of more diverse long-lived pionneers, recruited otherwise
below the intensity threshold. The resilience of tropical forests proved
consistent although several decades-long. Still, the disturbance impact
on communities redundancy cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Herault2018}.
